---
title: "Tomicro2024_ITS_Pre-processing"
author: "Theresa Kuhl-Nagel"
date: "2024-06-12"
output: html_document
---

```{r}
library(ggplot2) #ggplot2: Fancy graphs
library(ggpubr) #ggplot expansion for more fancy graphs
library(readxl)
library(readr)
library(dplyr)# Data handling
library(vegan)# Multivariate stat tool
library(tidyr) # Data handling
library(RColorBrewer) # Colours for fancy graphs
library(tibble)# Data handling: rownames/columnanmes
#install.packages("phyloseq")
#library(phyloseq) # tool for 16S amplicon analysis
library(lme4) # mixed effect models 
library(forcats)
library(tidyverse)
library(tibble)

```

Read data
```{r}
ASVs <- read.table("~/Megablast_Philemon_dada2_Unite_final.txt", header=T, sep="\t", dec=".", quote="", fill=FALSE) #add manually in textfile a 0 in front of single numbers 1, 2, 3, 4 ,5 ...
head(ASVs)
dim(ASVs)

tax <- read.table("~/Megablast_Philemon_dada2_Unite_taxonomy.txt", header=TRUE, sep="\t") 
head(tax)
```

```{r}
#change rownames
    ASVs$ASVid <- paste0("ASV",1:nrow(tax))
    tax$ASVid <- paste0("ASV",1:nrow(tax))

#separate taxonomy
tax <- tax %>%
      mutate(taxonomy = str_replace_all(taxonomy, "^(.*\\|)", "")) %>%
      mutate(taxonomy = str_replace_all(taxonomy, "k__", "")) %>%
      mutate(taxonomy = str_replace_all(taxonomy, "p__", "")) %>%
      mutate(taxonomy = str_replace_all(taxonomy, "c__", "")) %>%
      mutate(taxonomy = str_replace_all(taxonomy, "o__", "")) %>%
      mutate(taxonomy = str_replace_all(taxonomy, "f__", "")) %>%
      mutate(taxonomy = str_replace_all(taxonomy, "g__", "")) %>%
      mutate(taxonomy = str_replace_all(taxonomy, "s__", "")) %>%
      separate(taxonomy,
          into=c("kingdom", "phylum", "class", "order", "family", "genus", "species"), sep=";")

#Combine with ASV table
tax_ASV <- inner_join(ASVs, tax, by="ASVid")
tax_ASV
```

Filter out mitos and chloros and kingdoms other than fungi
```{r}
tax_ASV_filt <- tax_ASV%>%
  filter(kingdom == "Fungi")%>%
  filter(., !str_detect(order, "Mitochondria|Chloroplast"))%>%
  filter(., !str_detect(family, "Mitochondria|Chloroplast"))

tax_ASV_filt
dim(tax_ASV_filt)

#check if everything apart from Fungi is removed
unique(tax_ASV$kingdom)
unique(tax_ASV_filt$kingdom)

```

```{r}
#Remove taxonomy columns
ASV_subset <- tax_ASV_filt[,c(1:49)]
dim(ASV_subset)
head(ASV_subset)

#make first column rownames
rownames(ASV_subset) <- ASV_subset[, 1]
ASV_subset <- ASV_subset[, -1]

#put in correct order
ASV_subset <- ASV_subset[ , order(names(ASV_subset))]
View(ASV_subset)
```

Exclude ASVs with less than 5 hits in whole dataset
```{r}
ASVs_clean <- ASV_subset[-which(rowSums(ASV_subset[sapply(ASV_subset, is.numeric)]) <= 5), ]
dim(ASVs_clean)
head(ASVs_clean)

#Exclude ASVs with all 0 in one row
ASVs_no0 <- ASVs_clean[rowSums(ASVs_clean[, -1] > 0) != 0, ]
dim(ASVs_no0)
head(ASVs_no0)
View(ASVs_no0)

#Export
#write.table(ASVs_no0, "~/Megablast_Philemon_dada2_UNITE_no0_ASVNumbers2.txt", dec = ".", sep = "\t")

```

Clean taxonomy dataset
```{r}
# Convert row names to a column 
ASVs_no0_2 <- ASVs_no0 %>%
  rownames_to_column(var = "rowname")
head(ASVs_no0_2)
dim(ASVs_no0_2)

#change name of first column
colnames(tax_ASV_filt)[1] <- "rowname"
head(tax_ASV_filt)

#select columns in taxonomy
rowname <- tax_ASV_filt[, 1]
last_col <- tax_ASV_filt[, (ncol(tax_ASV_filt)-6):ncol(tax_ASV_filt)]
tax_new <- cbind(rowname, last_col)
head(tax_new)
dim(tax_new)

#join datasets
tax_clean <- inner_join(ASVs_no0_2, tax_new, by="rowname")
dim(tax_clean)
view(tax_clean)

#select taxonomy columns
ASVid <- tax_clean[, 1]
last_cols <- tax_clean[, (ncol(tax_clean)-6):ncol(tax_clean)]
tax_final <- cbind(ASVid, last_cols)
dim(tax_final)
head(tax_final)

#Export
#write.table(tax_final, "~/Megablast_Philemon_dada2_Unite_taxonomy_CLEAN.txt", dec = ".", sep = "\t")

```

#Create rarefied dataset
Read data
```{r}
ASVs_no0_new <- read.table("~/Megablast_Philemon_dada2_UNITE_no0_ASVNumbers2.txt", header=T, sep="\t", dec=".", fill=FALSE) 
head(ASVs_no0_new)
dim(ASVs_no0_new)
```

transpose data
```{r}
#transpose data
tASVs<- t(ASVs_no0_new) # samples in rows
dim(tASVs)
head(tASVs)
str(tASVs)
View(tASVs)
```

Visualization of rarefaction curve 1000 random rarefactions
```{r}
library(vegan)
myrare <- rarecurve(tASVs, step = 1000, cex=0.4)
```

get min rowSum 
```{r}
rowASVs <- rowSums(tASVs)
min(rowASVs)
```

creating 1000x rarefied table for further analysis of alpha-diversity
```{r}
Have_you_Runned_previously_RF<-"NO"
if (Have_you_Runned_previously_RF=="NO"){
  list_of_1000_rarefaction=lapply(1:1000, function(i){ rrarefy(tASVs, 18628)})
}

mean.dat <- as.data.frame(apply(simplify2array(lapply(list_of_1000_rarefaction, as.matrix)),1:2,mean))
view(mean.dat)
dim(mean.dat)

write.table(mean.dat, file="~/ASV_ITS_Rarefied_Table_1000_times_clean_ASVnumbers.txt", sep = "\t")

```
To account for uneven sequencing depth, we repeatedly performed 1000 rarefactions to our dataset’s lowest number of sequences. Average ASV abundances were calculated based on the 1000 rarefactions.



